Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.tex
Update results about Qs, not yet about thetas

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
\textbf{{[}Segway to WHAT TO MONITOR? Also, this paragraph is more or less the same as the third paragraph in the second section. Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}). These \(\theta\)s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\). A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient). Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. Therefore, this \(\theta\) is not sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

@@

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, folded, rank-normalized, split-chain \(\theta\)s. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

@@

We can interpret \(AC\)-values as a measure of stationarity. If \(AC\)-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative \(AC\)-values indicate divergence within imputation chains. \textbf{Subsequent sampled values within each imputation chain are less alike.} Positive \(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent iterations are similar, trending may occur. Negative \(AC\)-values show no threat to the stationarity component of convergence. On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence. As convergence diagnostic, the interest is therefore in positive \(AC\)-values. The magnitude of \(AC\)-values may be evaluated statistically {[}\textbf{add reference}{]}.

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms. But \(\widehat{R}=1\) and \(AC=0\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}. From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)). Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state. To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection. But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness proportion \(p_{mis}\) and a low number of iterations \(T\) will result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  \textbf{{[}Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas. The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.{]}}
\end{enumerate}

@@

Convergence of the MICE algorithm is investigated through model-based simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation set-up is summarized in the pseudo-code below. The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

Our results show the impact of inducing non-convergence in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics \(\widehat{R}\) and \(AC\) identify the signs of non-convergence in the algorithm. For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias. Bias in the descriptive statistics (\(Q=\mu\) and \(Q=\sigma\)) is most pronounced for the outcome variable \(Y\). Of the regression coefficients (\(Q=\beta\)) we observe the largest bias for \(\beta_1\) (the effect of \(X_1\) on \(Y\)). There is only one estimate for \(Q=r^2\) to consider. In the figures, we present simulation conditions where the number of iterations is \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \leq 30\). Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s. \textbf{{[}Add panel labels, describe panels{]}} As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\). The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\). The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\). This implies that, under the current specifications, no more than seven iterations are necessary to obtain optimal\textbf{/the best possible} performance in terms of bias and confidence validity.

@@

In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics \(\widehat{R}\) and \(AC\). \textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics indicate more signs of non-convergence in simulation conditions with a lower number of iterations. The missingness proportions, however, only seem to impact multivariate convergence. When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence, irrespective of the missingness proportion. When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.

>>>>

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
\textbf{{[}Segway to WHAT TO MONITOR? Also, this paragraph is more or less the same as the third paragraph in the second section. Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}). These \(\theta\)s may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\). A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient). Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. Therefore, this \(\theta\) is not sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

@@

\textbf{{[}Add: effective sample size is used if we want independent samples \emph{within} chains. In MI, we only use the final estimate. Also, computing n\_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand. But the default in iterative imp is often substantially lower.{]}}

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, rank-normalization, folding, and localization. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

@@

We can interpret \(AC\)-values as a measure of stationarity. If \(AC\)-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative \(AC\)-values indicate divergence within imputation chains. \textbf{Subsequent sampled values within each imputation chain are less alike.} Positive \(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent iterations are similar, trending may occur. Negative \(AC\)-values show no threat to the stationarity component of convergence. On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence. As convergence diagnostic, the interest is therefore in positive \(AC\)-values. The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}.

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms. But \(\widehat{R}\) and \(AC\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}. From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)). Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state. To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection. But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

@@

\textbf{In conclusion, we will use \(\widehat{R}\) conform \citet{veht19} and \(AC\) computed manually as non-convergence diagnostics in the simulation study. The diagnostics will be applied to four \(\theta\)s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness proportion \(p_{mis}\) and a low number of iterations \(T\) will result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  \textbf{{[}Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas. The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.{]}}
\end{enumerate}

@@

We investigate non-convergence of the MICE algorithm through model-based simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The number of simulation repetitions is 1000. The simulation set-up is summarized in the pseudo-code below. The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

Our results show the impact of inducing non-convergence in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics \(\widehat{R}\) and \(AC\) identify the signs of non-convergence in the algorithm. For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias. Bias in the descriptive statistics (\(Q=\mu\) and \(Q=\sigma\)) is most pronounced for the outcome variable \(Y\). Of the regression coefficients (\(Q=\beta\)) we observe the largest bias for \(\beta_1\) (the effect of \(X_1\) on \(Y\)). There is only one estimate for \(Q=r^2\) to consider. In the figures, we present simulation conditions where the number of iterations is \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \geq 30\). Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s. \textbf{{[}Add panel labels, describe panels{]}} As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\). The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\). The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\). This implies that, under the current specifications, no more than seven iterations are necessary to obtain \emph{sufficient} performance in terms of bias and confidence validity.

@@

In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics \(\widehat{R}\) and \(AC\). \textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations. The missingness proportions, however, only seem to impact multivariate convergence. When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence, irrespective of the missingness proportion. When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.
